Modelling BCG vaccination in the UK

### 2019-06-11 rmds/08-model-development.Rmd
added text to all super and sub scripts

The probability of transmission can be defined as the probability that a single contact between an infectious active TB case and a susceptible individual will lead to TB infection. The probability of transmission ($\beta_a$) can be redefined in terms of effective contacts ($C_{eff}$), historic effective contacts ($C^{hist}_{eff}$), actual average yearly total contacts ($C_{actual}$), the average period of time infectious ($\frac{1}{\nu_{avg}^{P}}$), and the average mortality rate ($\mu_{age}$) as follows,

\begin{equation}
\beta_a = \frac{(\nu_{avg}^{P} + \mu_{avg})\left(C_{eff} + C^{hist}_{eff} * t_{hist}\left(1 - \frac{log(t-1931)}{log(1990-1931)}\right)\right)}{C_{actual}}
(\#eq:beta-eq)
\end{equation}

@@

Vynnycky et al. found that the effective contact rate (i.e the transmission probability times the contact rate) for TB was approximately 22 in 1900 and fell to approximately 1 in 1990.[@Vynnycky1999] Incidence rates have increased since 1990 (Chapter \@ref(data)) and it is unclear what impact this has had on the effective contact rate. I have assumed that the effective contact rate is uniformly distributed between 0 and 5 contacts. Similarly, for the historic effective contact rate I have assumed a uniform distribution between 10 and 15. This gives a maximum effective contact rate of 20, inline with Vynnycky et al.'s estimate for the effective contact rate in 1900. I have also assumed that the historic contact rate declines over time using an exponential model, reducing to 0 in 1990. The prior for the half life of this decline ($c^{hist}_{half}$) has been assumed to be uniformly distributed between 0 and 20 years. Additional age stratification of $\beta$ is explored by including modifiers for certain age-groups. The baseline scenario is that no modification is required, with variation between children and adults ($\beta_{child}$) and between children, adults and older adults ($\beta_{child}$ and $\beta_{older adults}$) also being considered. The priors for these modifiers are normally distributed around 1 with a standard deviation of 0.5 (truncated to be greater than 0). The contact rate is estimated by averaging the total age-specific contact rates estimated from POLYMOD data (Section \@ref(contact-matrix)) on an annual basis.

@@

The time varying, age-specific, noisy, all-cause mortality rates ($\mu^{all-cause}_a(t)$) included in the demographic model are sourced from Office for National Statistics (ONS) estimates from 1981 until 2015. For years outside of the available data mortality rates are forecast using an age-stratified exponential model (\@ref(fig:mortality-england). This model was used as it constrains mortality rates above zero and decreases yearly changes in mortality rates over time. To model the uncertainty in the estimate of the annual number of deaths a normally distributed noise term was introduced with a standard deviation of 5%. In order to calculate the all-cause dynamic mortality rate ($\mu^{all-cause}_a(t)$), excluding deaths from, or related to, TB the following equation is used,

>>>>

The probability of transmission can be defined as the probability that a single contact between an infectious active TB case and a susceptible individual will lead to TB infection. The probability of transmission ($\beta_a$) can be redefined in terms of effective contacts ($C_{\text{eff}}$), historic effective contacts ($C^{\text{hist}}_{\text{eff}}$), actual average yearly total contacts ($C_{\text{actual}}$), the average period of time infectious ($\frac{1}{\nu_{\text{avg}}^{P}}$), and the average mortality rate ($\mu_{\text{age}}$) as follows,

\begin{equation}
\beta_a = \frac{(\nu_{\text{avg}}^{P} + \mu_{\text{avg}})\left(C_{\text{eff}} + C^{\text{hist}}_{\text{eff}} * t_{\text{hist}}\left(1 - \frac{log(t-1931)}{log(1990-1931)}\right)\right)}{C_{\text{actual}}}
(\#eq:beta-eq)
\end{equation}

@@

Vynnycky et al. found that the effective contact rate (i.e the transmission probability times the contact rate) for TB was approximately 22 in 1900 and fell to approximately 1 in 1990.[@Vynnycky1999] Incidence rates have increased since 1990 (Chapter \@ref(data)) and it is unclear what impact this has had on the effective contact rate. I have assumed that the effective contact rate is uniformly distributed between 0 and 5 contacts. Similarly, for the historic effective contact rate I have assumed a uniform distribution between 10 and 15. This gives a maximum effective contact rate of 20, inline with Vynnycky et al.'s estimate for the effective contact rate in 1900. I have also assumed that the historic contact rate declines over time using an exponential model, reducing to 0 in 1990. The prior for the half life of this decline ($c^{\text{hist}}_{\text{half}}$) has been assumed to be uniformly distributed between 0 and 20 years. Additional age stratification of $\beta$ is explored by including modifiers for certain age-groups. The baseline scenario is that no modification is required, with variation between children and adults ($\beta_{\text{child}}$) and between children, adults and older adults ($\beta_{\text{child}}$ and $\beta_{\text{older-adults}}$) also being considered. The priors for these modifiers are normally distributed around 1 with a standard deviation of 0.5 (truncated to be greater than 0). The contact rate is estimated by averaging the total age-specific contact rates estimated from POLYMOD data (Section \@ref(contact-matrix)) on an annual basis.

@@

The time varying, age-specific, noisy, all-cause mortality rates ($\mu^{\text{all-cause}}_a(t)$) included in the demographic model are sourced from Office for National Statistics (ONS) estimates from 1981 until 2015. For years outside of the available data mortality rates are forecast using an age-stratified exponential model (\@ref(fig:mortality-england). This model was used as it constrains mortality rates above zero and decreases yearly changes in mortality rates over time. To model the uncertainty in the estimate of the annual number of deaths a normally distributed noise term was introduced with a standard deviation of 5%. In order to calculate the all-cause dynamic mortality rate ($\mu^{\text{all-cause}}_a(t)$), excluding deaths from, or related to, TB the following equation is used,

### 2019-06-11 rmds/09-model-fitting.Rmd
added text to all super and sub scripts

$$O = \mathcal{N}\left(E_{syst}A, E_{noise}A\right)$$

Where $O$ are the observed notifications, $A$ are the incident cases of disease as forecast by the disease transmission model, $E_{syst}$ is the systematic reporting error, $E_{noise}$ is the reporting noise, and $\mathcal{N}$ represents the gaussian (normal) distribution. The priors for the model are defined in Table \@ref(tab:measurement-model). The prior for systematic reporting error is based on the assumption that cases are less likely to be reported than they are to be over reported. The prior for the reporting noise is an assumption. This observation model was also used for the non-UK born pulmonary cases.

@@

* Mixing rate between UK born and non-UK born ($M$).
* Effective contact rate ($c_{eff}$).
* Historic effective contact rate ($c^{hist}_{eff}$).
* Half life of the effective contact rate ($c^{hist}_{half}$).
* Systematic reporting error ($E_{syst}$).
* Reporting noise ($E_{noise}$).

@@

* Transmission probablity modifier for children ($\beta_{child}$).
* Transmission probablity modifier for older adults ($\beta_{older adults}$).

>>>>

$$O = \mathcal{N}\left(E_{\text{syst}}A, E_{\text{noise}}A\right)$$

Where $O$ are the observed notifications, $A$ are the incident cases of disease as forecast by the disease transmission model, $E_{\text{syst}}$ is the systematic reporting error, $E_{\text{noise}}$ is the reporting noise, and $\mathcal{N}$ represents the gaussian (normal) distribution. The priors for the model are defined in Table \@ref(tab:measurement-model). The prior for systematic reporting error is based on the assumption that cases are less likely to be reported than they are to be over reported. The prior for the reporting noise is an assumption. This observation model was also used for the non-UK born pulmonary cases.

@@

* Mixing rate between UK born and non-UK born ($M$).
* Effective contact rate ($c_{\text{eff}}$).
* The year that the current effective contact rate was reached ($c^{\text{year}}_{\text{eff}}$).
* Historic effective contact rate ($c^{\text{hist}}_{\text{eff}}$).
* Half life of the effective contact rate ($c^{\text{hist}}_{\text{half}}$).
* Systematic reporting error ($E_{\text{syst}}$).
* Reporting noise ($E_{\text{noise}}$).

@@

* Transmission probablity modifier for children ($\beta_{\text{child}}$).
* Transmission probablity modifier for older adults ($\beta_{\text{older-adults}}$).
